---
title: "Getting Started with delarr"
author: "delarr team"
date: "`r format(Sys.Date())`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with delarr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

`delarr` is a lightweight delayed matrix type for R designed for datasets that
are too large to fit comfortably in memory. Rather than loading an entire matrix
at once, `delarr` keeps data on disk (or in any external storage) and streams it
through your analysis pipeline in manageable chunks.
The package keeps its surface area deliberately small—one S3 class plus a
handful of verbs—while still supporting fused transformations, streaming
reductions, and pluggable backends. Operations are lazy: they build up a
pipeline of work that only executes when you call `collect()`. This lets you
chain multiple transformations without intermediate allocations.
This vignette walks through the essentials: creating delayed arrays, building
lazy pipelines, streaming to and from HDF5 files, and writing custom backends
for your own storage formats.

## Installation

`delarr` is under active development. Install the latest development version by
cloning the repository and loading it with `pkgload` or `devtools`:

```r
# install.packages("pkgload")
pkgload::load_all("/path/to/delarr")
```

Once installed, load the package:

```{r library, message=FALSE}
library(delarr)
```

## A first lazy pipeline

The core constructor `delarr()` wraps an existing matrix (or a seed backend).
All transformation verbs—`d_map()`, `d_center()`, `d_scale()`, etc.—return a new
`delarr` without actually computing anything. The operations are recorded and
fused together, then executed in a single pass when you call `collect()`.

This lazy evaluation has two major benefits:

1. **Memory efficiency**: Intermediate results are never materialised. A chain
   of ten transformations uses the same memory as one.
2. **Streaming**: When backed by on-disk storage, data flows through the
   pipeline in chunks, so you can process datasets larger than RAM.

```{r lazy-pipeline}
set.seed(1)
mat <- matrix(rnorm(20), 5, 4)
arr <- delarr(mat)

result <- arr |>
  d_center(dim = "rows", na.rm = FALSE) |>
  d_map(~ .x * 0.5) |>
  d_reduce(mean, dim = "rows") |>
  collect()

result
```

## Broadcasting and binary operations

Binary operations (`d_map2()` or arithmetic operators) remain lazy and support
broadcasting of scalars and row/column vectors. When you add a vector to a
matrix, `delarr` automatically determines whether it should broadcast across
rows or columns based on the vector length.
This makes common operations like subtracting row means or scaling by column
standard deviations simple and memory-efficient—the bias/scale vectors are
applied on-the-fly as each chunk streams through.

```{r broadcasting}
row_bias <- rnorm(nrow(mat))

delarr(mat) |>
  (`+`)(row_bias) |>
  (`*`)(1.5) |>
  d_reduce(mean, dim = "cols") |>
  collect(chunk_size = 2)
```

## Streaming to HDF5

HDF5 is a popular format for storing large numerical arrays on disk. The
`delarr_hdf5()` function creates a delayed array backed by an HDF5 dataset,
opening the file only when `collect()` executes and closing it immediately
after.

To write results back to disk without loading the full matrix into memory, pair
your pipeline with `hdf5_writer()`. This creates a streaming sink that receives
data chunk-by-chunk. The `compression` parameter controls gzip compression level
(0-9, with 4 as a good default balance of speed and size).

```{r hdf5, eval=knitr::is_html_output(), message=FALSE}
if (requireNamespace("hdf5r", quietly = TRUE)) {
  tf_in <- tempfile(fileext = ".h5")
  on.exit(unlink(tf_in), add = TRUE)
  input <- matrix(runif(30), 5, 6)
  f <- hdf5r::H5File$new(tf_in, mode = "w")
  f$create_dataset("X", robj = input)
  f$close_all()

  X <- delarr_hdf5(tf_in, "X")
  centred <- X |> d_center("cols")

  tf_out <- tempfile(fileext = ".h5")
  on.exit(unlink(tf_out), add = TRUE)
  writer <- hdf5_writer(tf_out, "X_centered", ncol = dim(centred)[2])
  collect(centred, into = writer, chunk_size = 3L)

  g <- hdf5r::H5File$new(tf_out, mode = "r")
  centred_back <- g[["X_centered"]]$read()
  g$close_all()

  centred_back
} else {
  "Install the 'hdf5r' package to run the streaming example."
}
```

## Custom backends with `delarr_seed()`

The real power of `delarr` comes from its pluggable backend system. Any storage
layer—databases, cloud storage, memory-mapped files, or custom binary
formats—can become a `delarr` backend by implementing a simple contract:

1. **Dimensions**: Know the total number of rows and columns

2. **Pull function**: Given row and column indices, return the corresponding
   submatrix

The `delarr_seed()` function wraps these two pieces into a seed object that
`delarr()` understands. The pull function is called lazily during `collect()`,
receiving only the indices needed for the current chunk.

```{r custom-backend}
random_backend <- list(
  pull = function(rows = NULL, cols = NULL) {
    rows <- if (is.null(rows)) seq_len(100) else rows
    cols <- if (is.null(cols)) seq_len(50) else cols
    matrix(rnorm(length(rows) * length(cols)), length(rows), length(cols))
  }
)

seed <- delarr_seed(
  nrow = 100,
  ncol = 50,
  pull = function(rows, cols) random_backend$pull(rows, cols)
)

rand_arr <- delarr(seed)
rand_arr |>
  d_map(~ .x^2) |>
  d_reduce(mean, dim = "cols") |>
  collect()
```

## Available verbs

`delarr` provides a focused set of transformation verbs:
- **`d_map()`** / **`d_map2()`**: Elementwise transformations using formula or
  function syntax
- **`d_center()`** / **`d_scale()`** / **`d_zscore()`**: Common preprocessing
  operations with row or column orientation
- **`d_detrend()`**: Remove linear trends from rows or columns
- **`d_where()`**: Masked updates—apply changes only where a condition holds
- **`d_reduce()`**: Row-wise or column-wise reductions (sum, mean, min, max, or
  custom functions)

All verbs return a new `delarr`, so you can chain them freely. The operations
fuse together and execute in a single streaming pass.

## When to use delarr
`delarr` is a good fit when:

- Your data is larger than available RAM
- You need to apply the same pipeline to many datasets
- You want streaming I/O without writing explicit chunking loops
- You prefer a simple, pipe-friendly API over S4 complexity

For sparse matrices, GPU acceleration, or parallel distributed computing,
consider specialised packages. `delarr` focuses on doing one thing well:
efficient lazy operations on dense, out-of-core matrices.
